### 2024-06-05 paper.tex
mv Fig6 caption to main text

We have implemented the methods just described in a Python package with a command line interface, \textit{MethSCAn} (Single-Cell ANalysis of METHylation data), which also offers a number of other functionalities for the analysis of scBS data. 
Fig.~\ref{fig:workflow} illustrates a typical scBS data analysis with \texttt{MethSCAn} and provides an overview of the implemented core functionalities.
The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}, which are first stored in a compressed format that enables efficient access to all CpG sites of the genome (see Methods).
A tutorial that showcases the analysis of a small example data set can be found at \url{https://anders-biostat.github.io/MethSCAn}.

@@

\begin{figure*}[p]
	\begin{center}
		\includegraphics[width=.8\textwidth]{figures/Fig_workflow.pdf}
	\end{center}
	\caption{\small \textbf{Overview of the functionalities implemented in the \textit{MethSCAn} package.}\\
		Common bisulfite sequencing mappers produce one methylation report file per cell.
		As it is inconvenient to work with thousands of text files, \texttt{methscan prepare} first stores their contents in a compressed format optimized for efficient access to genomic intervals.
		This command also produces cell-wise summary statistics that can be used to discard low-quality cells with \texttt{methscan filter}.
		The quality of each cell can furthermore be assessed by visualizing the average methylation around transcription start sites, which are known to show a characteristic dip in mammalian genomes, using \texttt{methscan profile}.
		This command can also visualize methylation at other genomic features of interest, such as enhancers or transcription factor binding sites.
		To obtain a methylation matrix analogous to scRNA-seq count matrices, the user must first decide in which genomic intervals methylation should be quantified.
		The user may either provide genomic regions of \emph{a priori} interest as a BED file, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{methscan scan}.
		The \texttt{methscan matrix} command then quantifies methylation at these genomic intervals in all cells.
		It provides two metrics:
		the simple percentage of methylated CpG sites and our proposed measure, the shrunken mean of the residuals, which is more robust to variation in sequencing coverage and stochastic differences in read position between cells.
		The resulting methylation matrix can then serve as input for established methods used for the exploration of single-cell data, such as dimensionality reduction and cell clustering.
		Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR (differentially methylated region) detection with \texttt{methscan diff}.
		This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
		These DMRs may then be associated with nearby genes, or used for GO enrichment with tools such as GREAT \citep{mclean2010great} to enable functional interpretation.
	}
	\label{fig:workflow}
\end{figure*}

>>>>

We have implemented the methods just described in a Python package with a command line interface, \textit{MethSCAn} (Single-Cell ANalysis of METHylation data), which also offers a number of other functionalities for the analysis of scBS data. 
Fig.~\ref{fig:workflow} illustrates a typical scBS data analysis with \texttt{MethSCAn} and provides an overview of the implemented core functionalities.
A tutorial that showcases the analysis of a small example data set can be found at \url{https://anders-biostat.github.io/MethSCAn}.

The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}.
Since it is inconvenient to work with hundreds or thousands of text files, \textit{MethSCAn} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome (see Methods).
\texttt{methscan prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.e.\ the number of CpG sites that have sequencing coverage.
These summary statistics can be used to detect cells with poor quality.
The quality of each single cell methylome can furthermore be inspected with \texttt{methscan profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS) at single-cell resolution.
The TSS profile is a useful quality control plot since methylation shows a characteristic dip roughly \textpm1~kb around the transcription start site in mammalian genomes.
Cells that do not show this pattern, or cells with few observed CpG sites, may then be discarded with \texttt{methscan filter}.

After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{methscan scan} command.
This produces a BED file that lists the genome coordinates of VMRs.
To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{methscan matrix}, which quantifies methylation at genomic intervals in all cells.
The command produces both, the simple percentage of methylated CpG sites, as well as our proposed methylation measure, the shrunken mean of the residuals, which is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any set of genomic features of interest, such as promoters, enhancers or transcription factor binding sites, obtaining one profile plot per cell.
The methylation matrix produced by the \texttt{matrix} function can then serve as input for established methods used for the exploration of single-cell data, such as dimensionality reduction and cell clustering.

Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR detection with \texttt{methscan diff}.
This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
These DMRs may then be associated with nearby genes, or used for GO enrichment with tools such as GREAT \citep{mclean2010great} to enable functional interpretation.

@@

\begin{figure*}[p]
	\begin{center}
		\includegraphics[width=.8\textwidth]{figures/Fig_workflow.pdf}
	\end{center}
	\caption{\small \textbf{Overview of the functionalities implemented in the \textit{MethSCAn} package.}\\
		Depicted is an scBS data analysis workflow that starts from methylation report files generated by mappers designed for bisulfite-converted sequencing reads.
		\textit{MethSCAn} facilitates exploration of this data by constructing a methylation matrix suitable for dimensionality reduction, and by enabling the detection of DMRs between groups of cells.
	}
	\label{fig:workflow}
\end{figure*}
